---
title: "Sub_metering Analytics Project"
author: "Itoro_E"
date: "7/2/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project centers around visualizations and time series analysis of sub-meters.

```{r}
# Load packages
#install.packages("RMySQL")
#install.packages("magrittr")
library(magrittr)
```
```{r}
# library(RMySQL)
# library(DBI)
library(pacman)
library(Hmisc)      #for descriptive statistics
library(tidyverse)  #Package for tidying data
library(lubridate)  #For working with dates/times of a time series
library(VIM)        #Visualizing and imputing missing values
library(broom)      #Tidy statistical summary output
library(knitr)      #report generation
library(kableExtra) #fancy table generator
library(psych)
library(lattice)
library(survival)
library(Formula)
```

Create a database connection
```{r}
 con = dbConnect(MySQL(), user='deepAnalytics', password='Sqltask1234!',
                 dbname='dataanalytics2018', host='data-analytics-2018.cbrosir2cswx.us-east-1.rds.amazonaws.com')
```

Use RMySQL to connect to the database and see what it contains.
List the tables contained in the database
```{r}
dbListTables(con)
```

Use the dbGetQuery function to download tables 2006 through 2010 with the specified attributes
```{r}
# yr_2006ALL <- dbGetQuery(con, "SELECT * FROM yr_2006")
# yr_2007ALL <- dbGetQuery(con, "SELECT * FROM yr_2007")
# yr_2008ALL <- dbGetQuery(con, "SELECT * FROM yr_2008")
# yr_2009ALL <- dbGetQuery(con, "SELECT * FROM yr_2009")
# yr_2010ALL <- dbGetQuery(con, "SELECT * FROM yr_2010")
```

create a Multi-Year data frame to serve as the primary data frame for the project.
using dplyr package function “bind_rows
Combine tables or df (ONLY includes the df that span an entire year: 2007, 2008, 2009)
```{r}
# newDF <- dplyr::bind_rows(yr_2007ALL, yr_2008ALL, yr_2009ALL)
```

Gather summary statistics
mean, mode, standard deviation, quartiles & characterization of the distribution
```{r}
# rm(M1_2007_month)
```

```{r}
# Summary statistics for newDF data features
# summary(newDF)
```

data munging newDF to create a 'DateTime' attribute Combining the 'Date' and 'Time' col within df
```{r}
# Combining the Date and Time columns with 'cbind' function in dplyr 
# newDF1 <-cbind(newDF,paste(newDF$Date,newDF$Time), stringsAsFactors=FALSE)
```

Header name for new attribute in the 11th column
```{r}
# colnames(newDF1)[11] <-"DateTime"
```

Move the DateTime attribute within the dataset to make it first column
```{r}
# newDF2 <- newDF1[,c(ncol(newDF1), 1:(ncol(newDF1)-1))]
```

Using "SELECT" function to DROP old "Date" and "Time" columns from newDF2
```{r}
newDF3 <- dplyr::select(newDF2, -c(Date,Time, id))
```

```{r}
# Confirm class of new DateTime feature is converted
# class(newDF3$DateTime)
```

Convert data type of new DateTime feature
```{r}
#newDF3$DateTime <- as.POSIXct(newDF3$DateTime, "%Y/%m/%d %H:%M:%S")
```

```{r}
newDF3$DateTime <- as.POSIXct(newDF3$DateTime, format = "%Y%m%d %H:%M:%S")

```

```{r}
attr(newDF3$DateTime, "tzone") <- "America/Chicago"
```

Using Libridate to create attributes for quarter, month, week, weekday,# day, hour and minute
```{r}
newDF3$year <- year(newDF3$DateTime)
newDF3$quarter <- quarter(newDF3$DateTime)
newDF3$month <- month(newDF3$DateTime)
newDF3$week <- week(newDF3$DateTime)
newDF3$day <- day(newDF3$DateTime)
newDF3$hour <- hour(newDF3$DateTime)
newDF3$minute <- minute(newDF3$DateTime)
```

Granularity - Subsetting and Meaningful Time Periods
```{r}
## Subset the second week of 2008 - All Observations
Week_2008 <- filter(newDF3, year == 2008 & week == 2)

```

```{r}
## Plot subset houseWeek of 2008
plot(Week_2008$Sub_metering_1)
```

```{r}
# mean(is.na(newDF3))
```

```{r}
# Subset the 9th day of January 2008 - All observations
House_9thDay <- filter(newDF3, year == 2008 & month == 1 & day == 9)
```

```{r}
# Plot sub-meter 1
plot_ly(House_9thDay, x = ~House_9thDay$DateTime, y = ~House_9thDay$Sub_metering_1, type = 'scatter', mode = 'lines')
```

```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - All observations 
plot_ly(House_9thDay, x = ~House_9thDay$DateTime, y = ~House_9thDay$Sub_metering_1, name = 'Kitchen',
        type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~House_9thDay$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~House_9thDay$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption January 9th, 2008",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))

```

```{r}
## Subset the 9th day of January 2008 - 10 Minute frequency
House_10min <- filter(newDF3, year == 2008 & month == 1 & day == 9 & (minute == 0 | minute == 10 | minute == 20 | minute == 30 | minute == 40 | minute == 50))
```

```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - 10 Minute frequency
plot_ly(House_10min, x = ~House_10min$DateTime, y = ~House_10min$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~House_10min$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~House_10min$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption January 9th, 2008",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))

```

With the granularity adjusted we get a much more clear picture of the power consumption on January 9th.
So what can you learn from this visualization? Your analysis could include answers to the following.

What peaks might represent the water heater?
Water might be used between 6:30 AM and 8:30 AM for bath. And about 7:30 AM and 2:30 PM
Water heater is also peaked during laundry

How about the AC? 
The spike in energy usage may not be realated to AC usage, since this the month of January/winter in Paris

What could be happening in the laundry room? 
Energy usage seems to peak every 2 to 2 1/2 hours from 2 AM to 10 PM
There seems to be a energy regulator device or appliance in the laundry area

How many times during this day are kitchen appliances being used? 
The homeowner uses more kitchen appliances between 5 PM and 6:30 PM

Does the data from these three sub-meters contain useful information for the homeowner? 
The information in these data is limited. cannot make decision based on it.

```{r}
## Subset the 9th day of January 2008 - 30min frequency
House_30min <- filter(newDF3, year == 2008 & month == 1 & day == 9 & (minute == 0 | minute == 30 | minute == 60))
```

```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - 30min frequency
plot_ly(House_30min, x = ~House_30min$DateTime, y = ~House_30min$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~House_30min$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~House_30min$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption January 9th, 2008",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))

```

```{r}
## Subset the 9th day of January 2009 - 10min frequency to See if Different from same day 2008
House_10min09 <- filter(newDF3, year == 2009 & month == 1 & day == 9 & (minute == 0 | minute == 10 | minute == 20 | minute == 30 | minute == 40 | minute == 50))
```

```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - 10min frequency for 2009
plot_ly(House_10min09, x = ~House_10min09$DateTime, y = ~House_10min09$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~House_10min09$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~House_10min09$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption January 9th, 2009",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))

```

2.1 Creating visualization with plotly for a random "Week" and "Day". Using all three sub-meters. Experiment with granularity. 
```{r}
## Subset the 28 week of 2009 - All Observations
Week_2009 <- filter(newDF3, year == 2009 & week == 28)
```

```{r}
# Subset the 5th day of July 2009 - All observations
House_July5 <- filter(newDF3, year == 2009 & month == 7 & day == 5)
```

```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - All observations 
plot_ly(House_July5, x = ~House_July5$DateTime, y = ~House_July5$Sub_metering_1, name = 'Kitchen',
        type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~House_July5$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~House_July5$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption July 9th, 2009",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))
```

Looking with the Granularity adjustment for clearer picture
```{r}
## Subset the 9th day of July 2009 - 10min frequency
House_10min_July <- filter(newDF3, year == 2009 & month == 7 & day == 5 & (minute == 0 | minute == 10 | minute == 20 | minute == 30 | minute == 40 | minute == 50))
```

```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - 10min frequency for 2009
plot_ly(House_10min_July, x = ~House_10min_July$DateTime, y = ~House_10min_July$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~House_10min_July$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~House_10min_July$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption July 5th, 2009",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))
```


3. Visualization for a time period. Both "Day" and "Week" highlighting typical patterns in a home. Another period of time that could provide insights using plotly. Experimenting with granularity until a visualization that maximizes information gain. 

```{r}
## Subset the week 45 of 2007 - All Observations
Week_2007 <- filter(newDF3, year == 2007 & week == 45)
```

```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - All observations for Week 45 Year 2007
plot_ly(Week_2007, x = ~Week_2007$DateTime, y = ~Week_2007$Sub_metering_1, name = 'Kitchen',
        type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~Week_2007$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~Week_2007$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption January 9th, 2008",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))

```

```{r}
# Subset the 5th day of November 2007 - All observations
House_Nov5 <- filter(newDF3, year == 2007 & month == 11 & day == 5)
```

```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - All observations 
plot_ly(House_9thDay, x = ~House_9thDay$DateTime, y = ~House_9thDay$Sub_metering_1, name = 'Kitchen',
        type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~House_9thDay$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~House_9thDay$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption January 9th, 2008",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))

```

With Granularity adjustment
```{r}
## Subset the 9th day of July 2009 - 10min frequency
House_10min_July <- filter(newDF3, year == 2009 & month == 7 & day == 5 & (minute == 0 | minute == 10 | minute == 20 | minute == 30 | minute == 40 | minute == 50))
```


```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - All observations 
plot_ly(House_9thDay, x = ~House_9thDay$DateTime, y = ~House_9thDay$Sub_metering_1, name = 'Kitchen',
        type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~House_9thDay$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~House_9thDay$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption January 9th, 2008",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))

```







```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - All observations 
plot_ly(House_July5, x = ~House_July5$DateTime, y = ~House_July5$Sub_metering_1, name = 'Kitchen',
        type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~House_July5$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~House_July5$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption July 9th, 2009",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))
```

Looking with the Granularity adjustment for clearer picture
```{r}
## Subset the 9th day of July 2009 - 10min frequency
House_10min_July <- filter(newDF3, year == 2009 & month == 7 & day == 5 & (minute == 0 | minute == 10 | minute == 20 | minute == 30 | minute == 40 | minute == 50))
```

```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - 10min frequency for 2009
plot_ly(House_10min_July, x = ~House_10min_July$DateTime, y = ~House_10min_July$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~House_10min_July$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~House_10min_July$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption July 5th, 2009",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))
```
```{r}
#rm(House_July9)
```

To aid with visualization of the three sub-meters data on the same chart
new column ‘Sub_Meter’ which contains all three sub-meters is created.
Another column ‘Watt_hr’ is created to store observed values for the sub-meters 
```{r}
# newDF_Data <- newDF3 %>%
#   gather(Meter, Watt_hr, 'Sub_metering_1', 'Sub_metering_2', 'Sub_metering_3')
```

Converting 'Meter' feature to categorical factor
```{r}
#newDF_Data$Meter <- factor(newDF_Data$Meter)
```

Use glimpse() to check on the Meter conversion before exploratory data analysis
```{r}
#glimpse(newDF_Data)
```

Data Exploration and Visualizations of Energy Usage Across a time periods in Sub-Meters
```{r}
#-Year_Proportional Plot
# newDF_Data %>%
#   group_by(year(DateTime), Meter) %>%
#   summarise(sum=sum(Watt_hr)) %>%
#   ggplot(aes(x=factor(`year(DateTime)`), sum, group=Meter,fill=Meter)) +
#   labs(x='Year', y='Proportion of Usage') +
#   ggtitle('Proportion of Sum of Yearly Energy Consumption') +
#   geom_bar(stat='identity', position='fill', color='green') +
#   theme(panel.border=element_rect(colour='green', fill=NA)) +
#   theme(text = element_text(size = 14))
```

```{r}
# by_Quarterly bar plot
# newDF_Data %>%
#   filter(year(DateTime)<2010) %>%
#   group_by(quarter(DateTime), Meter) %>%
#   summarise(sum=round(sum(Watt_hr/1000),3)) %>%
#   ggplot(aes(x=factor(`quarter(DateTime)`), y=sum)) +
#   labs(x='Quarter of the Year', y='kWh') +
#   ggtitle('Sum of Quarterly Energy Consumption') +
#   geom_bar(stat='identity', aes(fill = Meter), color='green') +
#   theme(panel.border=element_rect(colour='green', fill=NA)) +
#   theme(text = element_text(size = 14))
```

```{r}
# by_Month bar chart
# newDF_Data %>%
#   filter(year(DateTime)<2010) %>%
#   mutate(Month=lubridate::month(DateTime, label=TRUE, abbr=TRUE)) %>%
#   group_by(Month, Meter) %>%
#   summarise(sum=round(sum(Watt_hr)/1000),3) %>%
#   ggplot(aes(x=factor(Month), y=sum)) +
#     labs(x='Month of the Year', y='kWh') +
#     ggtitle('Monthly Sum of Energy Usage in the Year') +
#     geom_bar(stat='identity', aes(fill = Meter), colour='green') +
#   theme(panel.border=element_rect(colour='green', fill=NA)) +
#   theme(text = element_text(size = 14))
```

